\chapter{Conclusion}
\label{chap:conclusion}

% Summary of what the project achieved relative to its stated objectives
\section{Achievements}
\label{sec:achievements}

The following Achievements were achieved in this project:

\begin{itemize}
\item
  Data collection and vehicle allocation from relevant sources.
\item
  Configuration of SUMO for the Thapathali intersection, including road
  layout,traffic flow, and signal configuration.
\item
  Setup of the simulation environment for testing the traffic control
  system.
\item
  Fixed-timer logic configuration for baseline comparison.
\item
  DQN agent design and implementation for traffic signal optimisation.
\item
  The average waiting time over the 16-hour period (6 AM to 10 PM) was
  reduced by 22.90\%.
\item
  The average queue length over the 16-hour period was reduced by
  30.06\%.
\item
  The average throughput per waiting second over the 16-hour period
  improved by 32.65\%.
\end{itemize}

% Constraints, shortcomings, and areas where further work is warranted
\section{Limitations}
\label{sec:limitations}

Some factors were not considered or implemented in the current system;
these are outlined below as limitations.

\begin{itemize}
\item
  No hardware implementation or real-world field testing with traffic
  signals.
\item
  Pedestrians and public transport are excluded in this phase to keep
  the control problem
\item
  tractable and focus on vehicle-delay reduction at Thapathali; modeling
  them requires
\item
  additional constraints and reliable data.
\item
  Extreme scenarios or multi-modal integration, like ride-sharing or
  parking systems,
\item
  are not considered.
\item
  Random traffic signal timers were placed at the end of outgoing lanes
  to simulate network backspill; however, these were not based on the
  actual traffic conditions of the neighbouring network
\end{itemize}

% Directions for extending this work in scope, methodology, or deployment
\section{Future Work}
\label{sec:future_work}

Future development may include:

\begin{itemize}
\item
  Multi-Intersection Coordination: Scaling the system to manage multiple
  connected
\end{itemize}

\begin{quote}
intersections.
\end{quote}

\begin{itemize}
\item
  Advanced Optimization: Exploring hybrid algorithms or multi-agent RL
  for more
\end{itemize}

complex network topologies.

% Closing statement on the project's overall contribution and significance
\section{Final Remarks}
\label{sec:final_remarks}

This project set out to investigate whether a reinforcement
learning-based approach could meaningfully improve traffic signal
control at the Thapathali intersection in Kathmandu, one of the
city\textquotesingle s most persistently congested arterial junctions.
The results suggest that it can, and the case for adaptive signal
control is stronger than the numbers alone convey.

The DQN agent demonstrated consistent improvement over fixed-timer logic
across the full operational window. Where the gains are most evident is
during low-to-moderate demand periods, hours where fixed timing is
structurally wasteful, holding vehicles at red signals long after
opposing flows have cleared. In these windows, the
agent\textquotesingle s ability to read and react to real-time queue
states yielded noticeably better outcomes, highlighting just how rigid
and inefficient schedule-based control becomes when traffic is light and
variable.

The more telling story, however, lies in the peak hours. Improvements
during high-demand periods, the morning and evening surges that define
Kathmandu\textquotesingle s daily congestion profile, were comparatively
modest. This is not a failure of the algorithm; it is a reflection of a
harder problem. At genuine saturation, when every approach is
simultaneously overloaded, no signal timing strategy alone can conjure
capacity that does not exist. The Thapathali intersection, like many in
the Kathmandu valley, operates well beyond its designed throughput
during peak windows. The physical geometry, lane discipline, and
competing turning movements create a level of congestion that signal
optimisation can soften but cannot resolve. The DQN agent improves the
allocation of insufficient green time; it does not manufacture more road
space.

This distinction matters for how the results should be interpreted. The
system performs well precisely where fixed timers fail most obviously,
and it reaches its natural ceiling precisely where the underlying
infrastructure is the binding constraint. That ceiling is not an
algorithmic limitation but an infrastructural one, and it points toward
what complementary interventions would be needed alongside smarter
signal control to meaningfully address peak-hour gridlock.

As a proof of concept, this work establishes a credible and reproducible
foundation. The simulation pipeline, agent design, and evaluation
framework are all in place. What lies ahead, including real-world
hardware integration, inclusion of pedestrians and public transport, and
coordination across multiple connected intersections, are well-defined
extensions rather than open questions, each building directly on what
has been demonstrated here.

Urban traffic management in cities like Kathmandu has long depended on
static, schedule-driven systems that cannot adapt to the variability and
intensity of real-world demand. This project offers substantive early
evidence that adaptive, learning-based control can meaningfully close
that gap, not as a complete solution, but as an important and scalable
component of one.
